Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2015-12-04 paint-contour-tree.tex
typos corrected

% \Sesh{Almost everything below is new/reordered. Please read through.}
All previous algorithms begin by sorting (at least) the critical points. Can we beat this sorting bound for certain instances,
and can we characterize which inputs are hard? Intuitively, points that are incomparable in the contour tree
do not need to be compared. Look at \Fig{relative} to see such an example. All previous algorithms waste time
sorting all the maxima. Also consider the surface of \Fig{sorting}. The final contour tree is basically two binary
trees joined at their roots, and we do not need to entire sorted order of critical points
to construct the contour tree.

@@

The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
has strong optimality properties. For convenience, fix some value of $t$, and consider the set
of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
from $t$ to $t\log t$. Consider some $C \in [t,t\log t]$, and consider the set of terrains
where the algorithm makes $C$ comparisons. Then \emph{any algorithm} must make roughly $C$
comparisons in the worst-case over this set. (All further details are 
in Appendix~\Sec{lb}.)

@@

\InNotSoCGVer{
The main theorem of this section asserts that contour trees of minimum dominant manifolds have a simple description.
The exact statement will require some definitions and notation.
We require the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
Conventionally, all edges are directed from higher to lower function value. 
}
\InSoCGVer{
The main theorem of this section asserts that contour trees of minimum dominant manifolds have a simple description. 
Intuitively, the cutting procedure of \Sec{rain} introduced ``stumps'' where we made cuts, which 
is why we allowed for non-dominant minima and maxima in the definition of extremum dominant manifolds.  
The punchline is that, ignoring these stumps, the contour tree of an extremum dominant manifold is equivalent to its \emph{join} (or \emph{split}) tree, defined in~\cite{csa-cctad-00}.
Hence the critical join tree below is just the join tree without these stumps.  
Precise definitions and proofs are given in \Sec{jstedm}.
}

@@

%\InNotSoCGVer{\algorithmRunTimeSection}
%%%%%%%%%%%%%%%%%%%%
\InSoCGVer{
\subsection{Upper Bounds for Running Time}\label{sec:runTimeSummary}

@@

Given the above technical definition, the proof of the following lemma is straightforward, though long and so has been moved to \Sec{runTime}.

@@

Our main result, \Thm{main-corr}, is an easy corollary of the above lemma. Specifically, consider a critical point $v$ of the initial input 
complex.  By \Thm{jordan} this vertex appears in exactly one of the pieces output by $\rain$.  As in the \Thm{main-corr} statement, let $\ell_v$ 
denote the length of the longest directed path passing through $v$ in the contour tree the input complex, and let $\ell_v'$ 
denote the longest directed path passing through $v$ in the join tree of the piece containing $v$.
By \Thm{surgery}, ignoring non-dominant extrema introduced from cutting (whose cost can be charged to a corresponding saddle), the join tree on each piece output by $\rain$ is isomorphic to some 
connected subgraph of the contour tree of the input complex, and hence $\ell_v'\leq \ell_v$.  
Moreover, $|H_v|$ only counts vertices in a $v$ to root path and so trivially $|H_v|\leq \ell_v'$, implying \Thm{main-corr}.

@@

At this point we can now see what the path decomposition referenced in theorem statement should be.  It is just the union of all the maximum 
path decomposition across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\jc(\MM_i))$.
Since all procedures besides computing the join trees take linear time in the size of the input complex, we can therefore 
compute the contour tree in time
\InSoCGVer{
\begin{align*}
\hspace{.8in} O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}& \\
=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}&
\end{align*}
}
\InNotSoCGVer{
\[
O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\jc(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}
=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
\]
}

@@

\newcommand{\lowerBoundbyPathDecomp}{
\section{Lower Bound by Path Decomposition}
\label{sec:lb}
We first prove a lower bound for join trees, and then generalize to contour trees.
Note that form of the theorem statements in this section differ from \Thm{main-lb}, 
as they are stated directly in terms of path decompositions.  \Thm{main-lb} 
is an immediate corollary of the final theorem of this section \Thm{path-main-lb}.

@@

%\algorithmRunTimeSection
\section{Proof of \Lem{runTimeUpper}}
\label{sec:runTime}
\proofofRunTime

>>>>

% \Sesh{Almost everything below is new/reordered. Please read through.}
All previous algorithms begin by sorting (at least) the critical points. Can we beat this sorting bound for certain instances,
and can we characterize which inputs are hard? Intuitively, points that are incomparable in the contour tree
do not need to be compared. Look at \Fig{relative} to see such an example. All previous algorithms waste time
sorting all the maxima. Also consider the surface of \Fig{sorting}. The final contour tree is basically two binary
trees joined at their roots, and we do not need the entire sorted order of critical points
to construct the contour tree.

@@

The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
has strong optimality properties. For convenience, fix some value of $t$, and consider the set
of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
from $t$ to $t\log t$. Consider some $C \in [t,t\log t]$, and consider the set of terrains
where the algorithm makes $C$ comparisons. Then \emph{any algorithm} must make roughly $C$
comparisons in the worst-case over this set. (All further details are in \InSoCGVer{Appendix~}\Sec{lb}.)

@@

\InNotSoCGVer{
The main theorem of this section asserts that contour trees of minimum dominant manifolds have a simple description.
The exact statement will require some definitions and notation.
We require the notions of \emph{join} and \emph{split} trees, as given by~\cite{csa-cctad-00}.
Conventionally, all edges are directed from higher to lower function value. 
}
\InSoCGVer{
The main theorem of this section asserts that contour trees of minimum dominant manifolds have a simple description. 
Intuitively, the cutting procedure of \Sec{rain} introduced ``stumps'' where we made cuts, which 
is why we allowed for non-dominant minima and maxima in the definition of extremum dominant manifolds.  
The punchline is that, ignoring these stumps, the contour tree of an extremum dominant manifold is equivalent to its \emph{join} (or \emph{split}) tree, defined in~\cite{csa-cctad-00}.
Hence the critical join tree below is just the join tree without these stumps.  
Additional definitions and proofs are given in \Sec{jstedm}.
}

@@

%\InNotSoCGVer{\algorithmRunTimeSection}
%%%%%%%%%%%%%%%%%%%%
\InSoCGVer{
\subsection{Upper Bounds for Running Time}\label{sec:runTime}

@@

Given the above technical definition, the proof of the following lemma is straightforward, though long and so has been moved to \Sec{runTimeProof}.

@@

Our main result, \Thm{main-corr}, is an easy corollary of the above lemma. Specifically, consider a critical point $v$ of the initial input 
complex.  By \Thm{jordan} this vertex appears in exactly one of the pieces output by $\rain$.  As in the \Thm{main-corr} statement, let $\ell_v$ 
denote the length of the longest directed path passing through $v$ in the contour tree of the input complex, and let $\ell_v'$ 
denote the longest directed path passing through $v$ in the join tree of the piece containing $v$.
By \Thm{surgery}, ignoring non-dominant extrema introduced from cutting (whose cost can be charged to a corresponding saddle), the join tree on each piece output by $\rain$ is isomorphic to some 
connected subgraph of the contour tree of the input complex, and hence $\ell_v'\leq \ell_v$.  
Moreover, $|H_v|$ only counts vertices in a $v$ to root path and so trivially $|H_v|\leq \ell_v'$, implying \Thm{main-corr}.

@@

At this point we can now see what the path decomposition referenced in theorem statement should be.  It is just the union of all the maximum 
path decomposition across the extremum dominant pieces, $\pmax(\cC(\MM)) = \cup_{i=1}^k \pmax(\cJ(\MM_i))$.
Since all procedures besides computing the join trees take linear time in the size of the input complex, we can therefore 
compute the contour tree in time
\InSoCGVer{
\begin{align*}
\hspace{.8in} O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\cJ(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}& \\
=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}&
\end{align*}
}
\InNotSoCGVer{
\[
O\pth{N+ \sum_{i=1}^k \pth{\sum_{p \in \pmax(\cJ(\MM_i))} |p|\log |p|} + t_i\alpha(t_i) + N_i}
=\break O\pth{\pth{\sum_{p \in \pmax(\cC(\MM))} |p|\log |p|} + t\alpha(t) + N}
\]
}

@@

\newcommand{\lowerBoundbyPathDecomp}{
\section{Lower Bound by Path Decomposition}
\label{sec:lb}
We first prove a lower bound for join trees, and then generalize to contour trees.
Note that the form of the theorem statements in this section differ from \Thm{main-lb}, 
as they are stated directly in terms of path decompositions.  \Thm{main-lb} 
is an immediate corollary of the final theorem of this section, \Thm{path-main-lb}.

@@

%\algorithmRunTimeSection
\section{Proof of \Lem{runTimeUpper}}
\label{sec:runTimeProof}
\proofofRunTime
